Refinement for open automata

### 2021-06-01 2021-Refinement/report.tex
Changes suggested by Rabea

\begin{exi}[Enable, variable-oriented]
The automaton drawn on the right side of Figure \ref{fig:enable} is \(\OA{\mbrc{»}}{»}{\mbrc{v}}{v \mapsto l}{\mbrc{l, r}}{T}\) where transitions in \(T\) are:
\begin{align*}
	\OT{»}{»}{x}{\mbrc{l \mapsto x}}{v = l \wedge x \neq \delta}{\mbrc{}} &&
	\OT{»}{»}{\tau}{\mbrc{l \mapsto \delta\mpar{x}, r \mapsto \act{acc}\mpar{y}}}{v = l \wedge x = y}{\mbrc{v \gets r}} &&
	\OT{»}{»}{x}{\mbrc{r \mapsto x}}{v = r}{\mbrc{}}
\end{align*}
It is an alternative implementation of the enable operator which is FH-bisimilar to the previous one, as proved in \marginpar{probably forte, check that}.
Note that in this example \(r\) and \(l\) are a hole names and also closed terms: this is a naming conflict, though it is not ambiguous because \(J \cap \rterms = \emptyset\) always holds.
The variables local to a transition are all the unbound variables in the expressions minus the automaton variables.
For instance \(v\) is a variable of the automaton (used in the guard).
It is not local to the transition.
In a run, when taking effective transitions (without local variables), its value is not substituted with a closed term.
For instance \nmm{\OT{»}{»}{\tau}{\mbrc{l \mapsto \tau}}{v = l \wedge \tau \neq \delta}{\mbrc{}}} is an effective transition generated from the second transition but not \nmm{\OT{»}{»}{\tau}{\mbrc{l \mapsto \tau}}{l = l \wedge \tau \neq \delta}{\mbrc{}}} (result of the forbidden substitution \(v \mapsto l\)).

@@

Composition is a complex process that generates big automata with complex and potentially simplifiable transitions.
However simplification is a hard problem that is not yet automatised on open automata, so from this point the guards and transitions will always be simplified by hand without further explaination.
An example of composition can be found in Appendix \ref{apx:composition}.
\begin{prop}[Parallel composition]
Let \(A_1\), \(A_2\), \(A_3\) be three open automata with respectively \(J_1\), \(J_2\) and \(J_3\) as hole names.
We have
\begin{align*}
	A_1\subst{A_2}{j}\subst{A_3}{j'} & = A_1\subst{A_3}{j'}\subst{A_2}{j} & j, j' & \in J_1 \\
	\mpar{A_1\subst{A_2}{j_1}}\subst{A_3}{j_2} & = A_1\subst{A_2\subst{A_3}{j_2}}{j_1} & \mpar{j_1, j_2} & \in J_1 \times J_2
\end{align*}
Where the equality is modulo isomorphism (state renaming, associativity, commutativity).
\end{prop}
This property is induced by the composition on open pNets.
Intuitively it holds because the cartesian product of states, the union of sets, the union of maps and the conjunction on formulas (in the guards) are commutative and associative.
This allows us to define the parallel composition \(A\psubst{A_j^{j \in J'}}\) as the result of any order of (sequential) composition where each \(A_j\) is composed in the hole \(j\). % OR pcomp [[j |-> A_j]]

@@

The important properties we consider for the relation on states are being a simulation and the reduction of deadlocks as defined in the following definitions.
\begin{defi}[Pre-simulation]
For two open automata \(A_1 \defobject \OAg[1]\) and \(A_2 \defobject \OAg[2]\), a relation on configurations \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) is a pre-simulation if it satisfies both
\defitem Initial configurations are related: \(\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}\);
\defitem From related states, all out-transitions from \(A_1\) can be simulated in \(A_2\) and their target states are related:
\begin{multline*}
	\forall \mpar{s_1, s_2} \in S_1 \times S_2, \bigsymb{\forall} t_1 \defobject \OTx{1}{}{1}{1} \in \fOT{s_1}, \forall \sigma: V_1 \uplus V_2 \uplus \fvars{t_1} \to \values, \\
	\mpar{\sigma \vdash R\mpar{s_1, s_2} \wedge g_1} \implies \bigsymb{\exists} t_2 \defobject \OTx{2}{}{2}{2} \in \fOT{s_2}, \exists \nu: \fvars{t_2} \to \values, \\
	\sigma \uplus \nu \vdash \alpha_1 = \alpha_2 \wedge \bigwedge_\subbox{j \in J'_1 \cap J'_2} \beta_{1j} = \beta_{2j} \wedge g_2 \wedge R\mpar{s'_1, s'_2}\psubst{\psi_1 \uplus \psi_2}
\end{multline*}
\end{defi}
The last formula means that for every pair of related states under valuation of the automaton variables and every possible transitions (\(\vdash g_1\)) from the first automaton, there is a possible transition (\(\vdash g_2\)) such that the produced action matches (\(\alpha_1 = \alpha_2\)), the same holes' action (\(J'_1 \cap J'_2\)) matches (\(\beta_{1j} = \beta_{2j}\)) and the target states are related after variable update.
Apart from holes, this is a natural extension of the notion of simulation on LTS, which is the same definition without variables nor holes nor guards.
In FH-bisimulation, holes with the same name will receive the same open automaton when composed, and the resulting open automaton will still be equivalent.
Here we assume that one of the related open automata can be obtained by composing an open automaton in a hole, open automaton which can have holes itself.
This implies that there is no relation between their holes, for instance composing \(A_i\) in \(j_i\) (\(1 \leq i \leq n\)) of \(A_0\) gives the set of holes \(J \defobject \mpar{\biguplus_{0 \leq i \leq n} J_i} \setminus \mset{j_i}{1 \leq i \leq n} \) that is neither subset nor superset of \(J_0\).
Yet we still assume that in holes with the same name, the same open automaton will be composed.
Hence only the holes in common have to match their actions.
However it is not sufficient to imply properties on the relation on automatons like being a preorder.
It is only a necessary condition for any relation that will be defined with ``simulation" in its name.

Deadlocks are reachable configurations and from which no further transition is possible.
They can arise unpredictably when parallelism is involved (which is the case in our composition) and they break the intended behaviour of a system.
By preventing them we could hope that the refinement preserves some liveness properties.
\begin{defi}[Deadlock reduction, intuitive definition]
A relation on configurations \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) is deadlock reducing if:
\defitem It is a pre-simulation;
\defitem If from related states, there is a possible transition in the specification, then there is a pair of matching transition:
\begin{multline*}
	\forall \mpar{s_1, s_2} \in S_1 \times S_2, \forall \sigma: V_1 \uplus V_2 \to \values, \mpar{\sigma \vdash R\mpar{s_1, s_2}} \implies \\
	\mpar{\exists t_2 \in \fOT{s_2}, \exists \nu: \fvars{t_2} \to \values, \sigma \uplus \nu \vdash \fguard{t_2}} \implies \\
	\bigsymb{\exists} \mpar{t_1, t_2} \defobject \mpar{\OTx{1}{}{1}{1}, \OTx{2}{}{2}{2}} \in \fOT{s_1} \times \fOT{s_2}, \exists \nu: \fvars{t_1} \uplus \fvars{t_2} \to \values, \\
	\sigma \uplus \nu \vdash g_1 \wedge g_2 \wedge \alpha_1 = \alpha_2 \wedge \bigwedge_\subbox{j \in J'_1 \cap J'_2} \beta_{1j} = \beta_{2j} \wedge R\mpar{s'_1, s'_2}\psubst{\psi_1 \uplus \psi_2}
\end{multline*}
\end{defi}
Actually, this requirement prevents the presence of deadlock state in the first automaton if there is a related non-deadlock state in the second by contraposition.
On top of that the transition preventing the deadlock has to be an existing transition in the specification.
This is because a valid transition unmatched in the specification is not a deadlock, yet we do not want it.
The direct formulation of the property would also have required the unintuitive quantification on a possibly empty set of transition to state that they are impossible.
The two previous reasons motivated the formulation in contraposed form.

``Deadlock reduction" conflicts with all the properties where composition is involved (for instance ``composition correctness") by disallowing some unwanted cases where an automaton in a hole cannot produce any action that any out-transition expects.
In particular filling a hole with the deadlock automaton (only one state without out-transitions) is not considered a refinement for a deadlock reducing relation.
A way to solve these conflicts is to characterise a composition that do not introduce deadlocks, which will be used instead of the composition introduced earlier.
\begin{defi}[Reachability]
For any open automata \(A \defobject \OAg\), a reachability predicate \(\reach{A}: S \to \rformulas[V]\) is a predicate on states satisfying both
\defitem Inital state is reachable: \(\sigma_0 \vdash \reach{A}\mpar{s_0}\)
\defitem Reachability is preserved across transitions: \nmm{\bigsymb{\forall} t \defobject \OTg \in T, \forall \nu: \fvars{t} \to \values, \nu \vdash \reach{A}\mpar{s} \wedge g \implies \reach{A}\mpar{s'}\psubst{\psi}}
\end{defi}
The reachability predicate is used to characterise the reachable configurations in a run of an automaton.
In fact the role of the predicate is to characterise potentially reachable configurations without having to characterise configurations as the result of a valid path in an automaton.

To do that we impose that the initial configuration is reachable and that reachability is preserved by taking valid transitions.
This effectively makes reachability account for all paths and potentially over-approximate the reachable configurations.
The exact reachability may not be reprensentable in the formulas, hence the need of potentially over-approximating. % TODO: is representable the correct word

@@

In formula, for \(A = \OAg\), it gives
\begin{multline*}
	\forall s \defobject \mpar{s_p, s_c} \in S, \forall \sigma: V \to \values, \mpar{\sigma \vdash \reach{A_p\subst{A_c}{j}}\mpar{s}} \implies \\
	\mpar{\exists t_p \in \fOT{s_p}, \exists \nu_p: \fvars{t_p} \to \values, \sigma \uplus \nu_p \vdash \fguard{t_p}} \implies \\
	\exists t \in \fOT{s}, \exists \nu: \fvars{t} \to \values, \sigma \uplus \nu \vdash \fguard{t}
\end{multline*}
\end{defi}
We can expand the aliases \(S, V\) into \(S_p \times S_c, V_p \uplus V_c\) and transform the last transition \(t\) into a pair \(\mpar{t_p, t}\) where the first transition is the one from \(A_p\) which generated \(t\).
Doing that makes the definition become mildly similar with the deadlock reduction one. % OR slightly
It is not a coincidence because their goal is to caracterise the same kind of compatibility between automata.
Actually it is possible to simplify both definition and make them even more similar.
\begin{defi}[Deadlock reduction, working definition]
A relation on configurations \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) is deadlock reducing if it is a pre-simulation and it satisfies the following:
\begin{multline*}
	\forall \mpar{s_1, s_2} \in S_1 \times S_2, \forall \sigma: V_1 \uplus V_2 \uplus \biguplus_\subbox{t_2 \in \fOT{s_2}} \fvars{t_2} \to \values, \\
	\sigma \vdash \mpar{R\mpar{s_1, s_2} \wedge \bigvee_\subbox{t_2 \in \fOT{s_2}} \fguard{t_2} \implies \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1}}
\end{multline*}
\end{defi}
The requirement here is that from any related configurations, if there is a possible transition in the second automaton (whatever the value of the free variables) then there is a possible transition in the first automaton (the existential quantifier is implicit in the notation \(\vdash\)).
The transition are not matched because the relation on configurations is a pre-simulation and it is sufficient to ensure it as will be shown in the proof.
Apart from the matching transitions this is exactly the same explaination as the intuitive definition.
The expected advantage of this formulation is that a SMT should behave better without the quantifiers.
Indeed the big disjunctions are not quantifiers because in practise the out-transitions are finitary, and so is the expansion of the disjunction. % WHAT IS THE ISSUE HERE?
\begin{lem}
The intuitive and working definition of deadlock reduction are equivalent.
\end{lem}
\begin{defi}[Non-locking composition, working definition]
The composition \(A \defobject A_p\subst{A_c}{j}\) is a non-locking composition if:
\[ \forall s \defobject \mpar{s_p, s_c} \in S, \forall \sigma: V \uplus \biguplus_\subbox{t_p \in \fOT{s_p}} \fvars{t_p} \to \values, \sigma \vdash \reach{A}\mpar{s} \wedge \bigvee_\subbox{t_p \in \fOT{s_p}} \fguard{t_p} \implies \bigvee_\subbox{t \in \fOT{s}} \fguard{t} \]
\end{defi}
\begin{lem}
The intuitive and working definition of non-locking composition are equivalent.
\end{lem}
An illustration of the introduction of a deadlock by composition and the definition of non-locking composition is given in Appendix \ref{apx:lockcomp}.
Proof of both lemma are given in Appendix \ref{apx:lemeqd}.

This new definition of composition can replace the standard composition in the case where only one hole is filled, and also be extended to a parallel composition.
For the parallel composition, we may think that if the composition is non-locking, then every composition is non-locking for all composition order.
It is not true.
Actually even if the last composition of any order of composition is non-locking there is no guarantee that this order is non-locking at each step.
However the oppsite way is true even if uninformative: % WHAT IS THE ISSUE?
If there is an order of composition where every composition is non-locking then the last composition is non-locking, last composition resulting in the same automaton as the parallel composition.
So the condition is better checked after the parallel composition or the last individual composition.
\begin{exi}
\begin{figure}
\centering
\input{file_sync.tex}
\caption{A made-up protocol for file synchronisation}
\label{fig:pnls}
\end{figure}
The automaton on Figure \ref{fig:pnls} is a sample protocol for file synchronisation.
The two holes are \(net\) a network manager used to perform operations on the network, and \(hash\) a file integrity check used to generate a fingerprint.
The state \(i\) is the initial state.
In that state we want to synchronise the files before working, or work on the local version as a fallback.
If the preferred version is the remote one, we go in state \(d1\) asking the netwotk manager to fetch the lastest version, then ask the integrity checker to validate the fingerprint.
If the preferred version is the local one, we go in state \(d2\) asking for the integrity checker to generate a fingerprint for the local files, then ask the network manager to push them.
In fallback we use the local version.

We can debranch the network manager and the file integrity check by composing a deadlock automaton in their holes, but always both at the same time.
The reason for that is the parallel composition is non-locking, but any individual composition is locking because the state \(d1\) or \(d2\) (depending on the order) becomes a deadlock.
\end{exi}
From this point and in the previous definitions, composition will only refer to non-locking composition.

@@

\section{Refinement relations in restricted cases}\label{sec:prelref}
The main goal of this section is to explain the different aspects of refinement in restricted cases so that they are not all introduced at once in the real refinement relation.
In particular if we have a family of open automata \(A_i \defobject \OAg[i]\), \(0 \leq i \leq n\), let \(A \defobject \OAg\) be such that \(A = A_0\psubst{j_i \mapsto A_i \middle| 1 \leq i \leq n}\).
Recall that there is no relation between the holes of \(A\) and \(A_0\) because \(J = \mpar{\biguplus_{0 \leq i \leq n} J_i} \setminus \mset{j_i}{i \in \mdbrk{1; n}}\) is neither subset nor superset of \(J_0\).
So looking at restricted cases where holes are related in a specific manner can help understanding the general case.
Hopefully the intermediate refinement relation introduced can be used as simpler (and less expensive to compute) versions of the real refinement relation in the case the manipulated automata respect some constraints.

We restrict the holes of the related automata so that we can explain the different refinements without their interactions.
In the following definition, \(\triangle \in \mbrc{\subseteq, =, \supseteq}\).
\begin{defi}[Hole-\(\triangle\) simulation]
For two open automata \(A_1 \defobject \OAg[1]\) and \(A_2 \defobject \OAg[2]\), the relation on configurations \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) is a hole-\(\triangle\) simulation of \(A_1\) by \(A_2\) if:
\defitem \(J_1 \triangle J_2\)
\defitem \(\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}\)
\defitem \(\forall \mpar{s_1, s_2} \in S_1 \times S_2,\)\vspace{-8pt}
\noindent\begin{multline*}
	\mpar{\everymath{\displaystyle}\begin{array}{l}
		\bigsymb{\forall} t_1 \defobject \OTx{1}{}{1}{1} \in \fOT{s_1}, \bigsymb{\exists} \mpar{t_{2x} \defobject \OTx{2}{x}{2x}{2x} \in \fOT{s_2}}^{x \in X}, \\
		\quad \mpar{\forall x \in X, J'_{2x} \cap J_1 = J'_1 \cap J_2} \\[-10pt]
		\nwedge V_1 \uplus V_2 \uplus \fvars{t_1} \vdash R\mpar{s_1, s_2} \wedge g_1 \implies \operatorname*{\bigsymb{\bigvee}}_{x \in X} \mpar{\begin{array}{l}
			\alpha_1 = \alpha_{2x} \wedge \bigwedge_\subbox{j \in J'_{2x} \cap J_1} \beta_{1j} = \beta_{2xj} \\[12pt]
			\nwedge g_{2x} \wedge R\mpar{s'_1, s'_{2x}}\psubst{\psi_1 \uplus \psi_{2x}}
		\end{array}}
	\end{array}} \\
	\wedge V_1 \uplus V_2 \uplus \biguplus_\subbox{t_2 \in \fOT{s_2}} \fvars{t_2} \vdash R\mpar{s_1, s_2} \wedge \bigvee_\subbox{t_2 \in \fOT{s_2}} \fguard{t_2} \implies \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1}
\end{multline*}
\end{defi}
The third item of the definition has two parts:
The first part is the main requirement, a (not equivalent) reordering of the pre-simulation requirement so that it better fits SMT.
Instead of matching an effective transition to another effective transition as in the pre-simulation definition, we match an open transition to a family of covering open-transitions.
The second part is directly the deadlock reduction requirement.

@@

\[ \iff \]

@@

		\end{array}\!\!}

>>>>

\begin{exi}[Enable, variable-oriented]
The automaton drawn on the right side of Figure \ref{fig:enable} is \(\OA{\mbrc{»}}{»}{\mbrc{v}}{\mbrc{v \mapsto l}}{\mbrc{l, r}}{T}\) where transitions in \(T\) are:
\begin{align*}
	\OT{»}{»}{x}{\mbrc{l \mapsto x}}{v = l \wedge x \neq \delta}{\mbrc{}} &&
	\OT{»}{»}{\tau}{\mbrc{l \mapsto \delta\mpar{x}, r \mapsto \act{acc}\mpar{y}}}{v = l \wedge x = y}{\mbrc{v \gets r}} &&
	\OT{»}{»}{x}{\mbrc{r \mapsto x}}{v = r}{\mbrc{}}
\end{align*}
It is an alternative implementation of the enable operator which is FH-bisimilar to the previous one, as proved in \marginpar{probably forte, check that}.
Note that in this example \(r\) and \(l\) are a hole names and also closed terms: this is a naming conflict, though it is not ambiguous because \(J \cap \rterms = \emptyset\) always holds.
The variables local to a transition are all the unbound variables in the expressions minus the automaton variables.
For instance \(v\) is a variable of the automaton (used in the guard).
It is not local to the transition.
In a run, when taking effective transitions (without local variables), its value is not substituted with a closed term.
For instance \nmm{\OT{»}{»}{\tau}{\mbrc{l \mapsto \tau}}{v = l \wedge \tau \neq \delta}{\mbrc{}}} is an effective transition generated from the second transition but not \nmm{\OT{»}{»}{\tau}{\mbrc{l \mapsto \tau}}{l = l \wedge \tau \neq \delta}{\mbrc{}}} (result of the forbidden substitution \(v \mapsto l\)).

@@

Composition is a complex process that generates big automata with complex and potentially simplifiable transitions.
However simplification is a hard problem that is not yet automatised on open automata, so from this point the guards and transitions will always be simplified by hand without further explaination.
An example of composition can be found in Appendix \ref{apx:composition}.
\begin{prop}[Parallel composition]
Let \(A_1\), \(A_2\), \(A_3\) be three open automata with respectively \(J_1\), \(J_2\) and \(J_3\) as hole names.
We have
\begin{align*}
	A_1\subst{A_2}{j}\subst{A_3}{j'} & = A_1\subst{A_3}{j'}\subst{A_2}{j} & j, j' & \in J_1 \\
	\mpar{A_1\subst{A_2}{j_1}}\subst{A_3}{j_2} & = A_1\subst{A_2\subst{A_3}{j_2}}{j_1} & \mpar{j_1, j_2} & \in J_1 \times J_2
\end{align*}
Where the equality is modulo isomorphism (state renaming, associativity, commutativity).
\end{prop}
This property is induced by the composition on open pNets.
Intuitively it holds because the cartesian product of states, the union of sets, the union of maps and the conjunction on formulas (in the guards) are commutative and associative.
This allows us to define the simultaneous composition \(A\psubst{A_j^{j \in J'}}\) as the result of any order of (sequential) composition where each \(A_j\) is composed in the hole \(j\). % psubst to pcomp macro [[]]

@@

The important properties we consider for the relation on states are being a simulation and the prevention of deadlocks as defined in the following definitions.
\begin{defi}[Pre-simulation]
For two open automata \(A_1 \defobject \OAg[1]\) and \(A_2 \defobject \OAg[2]\), a relation on configurations \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) is a pre-simulation if it satisfies both
\defitem Initial configurations are related: \(\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}\);
\defitem From related states, all out-transitions from \(A_1\) can be simulated in \(A_2\) and their target states are related:
\begin{multline*}
	\forall \mpar{s_1, s_2} \in S_1 \times S_2, \bigsymb{\forall} t_1 \defobject \OTx{1}{}{1}{1} \in \fOT{s_1}, \forall \sigma: V_1 \uplus V_2 \uplus \fvars{t_1} \to \values, \\
	\mpar{\sigma \vdash R\mpar{s_1, s_2} \wedge g_1} \implies \bigsymb{\exists} t_2 \defobject \OTx{2}{}{2}{2} \in \fOT{s_2}, \exists \nu: \fvars{t_2} \to \values, \\
	\sigma \uplus \nu \vdash \alpha_1 = \alpha_2 \wedge \bigwedge_\subbox{j \in J'_1 \cap J'_2} \beta_{1j} = \beta_{2j} \wedge g_2 \wedge R\mpar{s'_1, s'_2}\psubst{\psi_1 \uplus \psi_2}
\end{multline*}
\end{defi}
The last formula means that for every pair of related configurations of the automaton and every possible transitions (\(\vdash g_1\)) from the first automaton, there is a possible transition (\(\vdash g_2\)) such that the produced action matches (\(\alpha_1 = \alpha_2\)), the same holes' action (\(J'_1 \cap J'_2\)) matches (\(\beta_{1j} = \beta_{2j}\)) and the target states are related after variable update. % OR variableS
Apart from holes, this is a natural extension of the notion of simulation on LTS, which is the same definition without variables nor holes nor guards.
In FH-bisimulation, holes with the same name will receive the same open automaton when composed, and the resulting open automaton will still be equivalent.
Here we assume that one of the related open automata can be obtained by composing an open automaton in a hole, open automaton which can have holes itself.
This implies that there is no relation between their holes, for instance composing \(A_i\) in \(j_i\) (\(1 \leq i \leq n\)) of \(A_0\) gives the set of holes \(J \defobject \mpar{\biguplus_{0 \leq i \leq n} J_i} \setminus \mset{j_i}{1 \leq i \leq n} \) that is neither subset nor superset of \(J_0\).
Yet we still assume that in holes with the same name, the same open automaton will be composed.
Hence only the holes in common have to match their actions.
However it is not sufficient to imply properties on the relation on automatons like being a preorder.
It is only a necessary condition for any relation that will be defined with ``simulation" in its name.

Deadlocks are reachable configurations from which no further transition is possible.
They can arise unpredictably when parallelism is involved (which is the case in our composition) and they break the intended behaviour of a system.
By preventing them we could hope that the refinement preserves some liveness properties.
\begin{defi}[Deadlock prevention, intuitive definition]
A relation on configurations \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) is deadlock reducing if:
\defitem It is a pre-simulation;
\defitem If from related states, there is a possible transition in the specification, then there is a pair of matching transition:
\begin{multline*}
	\forall \mpar{s_1, s_2} \in S_1 \times S_2, \forall \sigma: V_1 \uplus V_2 \to \values, \mpar{\sigma \vdash R\mpar{s_1, s_2}} \implies \\
	\mpar{\exists t_2 \in \fOT{s_2}, \exists \nu: \fvars{t_2} \to \values, \sigma \uplus \nu \vdash \fguard{t_2}} \implies \\
	\bigsymb{\exists} \mpar{t_1, t_2} \defobject \mpar{\OTx{1}{}{1}{1}, \OTx{2}{}{2}{2}} \in \fOT{s_1} \times \fOT{s_2}, \exists \nu: \fvars{t_1} \uplus \fvars{t_2} \to \values, \\
	\sigma \uplus \nu \vdash g_1 \wedge g_2 \wedge \alpha_1 = \alpha_2 \wedge \bigwedge_\subbox{j \in J'_1 \cap J'_2} \beta_{1j} = \beta_{2j} \wedge R\mpar{s'_1, s'_2}\psubst{\psi_1 \uplus \psi_2}
\end{multline*}
\end{defi}
Actually, this requirement prevents the presence of deadlock state in the first automaton if there is a related non-deadlock state in the second by contraposition.
On top of that the transition preventing the deadlock has to be an existing transition in the specification.
This is because a valid transition unmatched in the specification is not a deadlock, yet we do not want it.
The direct formulation of the property would also have required the unintuitive quantification on a possibly empty set of transition to state that they are impossible.
The two previous reasons motivated the formulation in contraposed form.

``Deadlock prevention" conflicts with all the properties where composition is involved (for instance ``composition correctness") by disallowing some unwanted cases where an automaton in a hole cannot produce any action that any out-transition expects.
In particular filling a hole with the deadlock automaton (only one state without out-transitions) is not considered a refinement for a deadlock reducing relation.
A way to solve these conflicts is to characterise a composition that do not introduce deadlocks, which will be used instead of the composition introduced earlier.
\begin{defi}[Reachability]
For any open automata \(A \defobject \OAg\), a reachability predicate \(\reach{A}: S \to \rformulas[V]\) is a predicate on states satisfying both
\defitem Inital state is reachable: \(\sigma_0 \vdash \reach{A}\mpar{s_0}\)
\defitem Reachability is preserved across transitions: \nmm{\bigsymb{\forall} t \defobject \OTg \in T, \forall \nu: \fvars{t} \to \values, \nu \vdash \reach{A}\mpar{s} \wedge g \implies \reach{A}\mpar{s'}\psubst{\psi}}
\end{defi}
The reachability predicate is used to characterise the reachable configurations in a run of an automaton.
In fact the role of the predicate is to characterise potentially reachable configurations without having to characterise configurations as the result of a valid path in an automaton.

To do that we impose that the initial configuration is reachable and that reachability is preserved by taking valid transitions.
This effectively makes reachability take into account all paths, and potentially over-approximate the reachable configurations.
The exact reachability may not be representable in the formulas, hence the need of potentially over-approximating.

@@

Formally for \(A = \OAg\), it gives
\begin{multline*}
	\forall s \defobject \mpar{s_p, s_c} \in S, \forall \sigma: V \to \values, \mpar{\sigma \vdash \reach{A_p\subst{A_c}{j}}\mpar{s}} \implies \\
	\mpar{\exists t_p \in \fOT{s_p}, \exists \nu_p: \fvars{t_p} \to \values, \sigma \uplus \nu_p \vdash \fguard{t_p}} \implies \\
	\exists t \in \fOT{s}, \exists \nu: \fvars{t} \to \values, \sigma \uplus \nu \vdash \fguard{t}
\end{multline*}
\end{defi}
We can expand the aliases \(S, V\) into \(S_p \times S_c, V_p \uplus V_c\) and transform the last transition \(t\) into a pair \(\mpar{t_p, t}\) where the first transition is the one from \(A_p\) which generated \(t\).
Doing that makes the definition become mildly similar with the deadlock prevention one. % OR slightly
It is not a coincidence because their goal is to caracterise the same kind of compatibility between automata.
Actually it is possible to simplify both definition and make them even more similar.
\begin{defi}[Deadlock prevention, working definition]
A relation on configurations \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) is deadlock reducing if it is a pre-simulation and it satisfies the following:
\begin{multline*}
	\forall \mpar{s_1, s_2} \in S_1 \times S_2, \forall \sigma: V_1 \uplus V_2 \uplus \biguplus_\subbox{t_2 \in \fOT{s_2}} \fvars{t_2} \to \values, \\
	\sigma \vdash \mpar{R\mpar{s_1, s_2} \wedge \bigvee_\subbox{t_2 \in \fOT{s_2}} \fguard{t_2} \implies \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1}}
\end{multline*}
\end{defi}
The requirement here is that from any related configurations, if there is a possible transition in the second automaton (whatever the value of the free variables) then there is a possible transition in the first automaton (the existential quantifier is implicit in the notation \(\vdash\)).
The transition are not matched because the relation on configurations is a pre-simulation and it is sufficient to ensure it as will be shown in the proof.
Apart from the matching transitions this is exactly the same explaination as the intuitive definition.
The expected advantage of this formulation is that a SMT should behave better without the quantifiers.
Indeed the big disjunctions are not quantifiers because in practise the out-transitions are finitary, and so is the expansion of the disjunction. % Check if that construction is correct
\begin{lem}
The intuitive and working definition of deadlock prevention are equivalent.
\end{lem}
\begin{defi}[Non-locking composition, working definition]
The composition \(A \defobject A_p\subst{A_c}{j}\) is a non-locking composition if:
\[ \forall s \defobject \mpar{s_p, s_c} \in S, \forall \sigma: V \uplus \biguplus_\subbox{t_p \in \fOT{s_p}} \fvars{t_p} \to \values, \sigma \vdash \reach{A}\mpar{s} \wedge \bigvee_\subbox{t_p \in \fOT{s_p}} \fguard{t_p} \implies \bigvee_\subbox{t \in \fOT{s}} \fguard{t} \]
\end{defi}
\begin{lem}
The intuitive and working definition of non-locking composition are equivalent.
\end{lem}
An illustration of both the introduction of a deadlock by composition and the definition of non-locking composition is given in Appendix \ref{apx:lockcomp}.
Proof of both lemma are given in Appendix \ref{apx:lemeqd}.

This new definition of composition can replace the standard composition in the case where only one hole is filled, and also be extended to a simultaneous composition.
For the simultaneous composition, we may think that if the last composition of a specific composition order is non-locking, then every composition is non-locking for all composition order. % orderS
Unfortunately it is not true.
Actually even if the last composition of any order of composition is non-locking there is no guarantee that this order is non-locking at each intermediate step.
However the opposite way is true even if uninformative:
If there is an order of composition where every composition is non-locking then the last composition is non-locking, last composition resulting in the same automaton as the simultaneous composition.
So the condition is better checked after the simultaneous composition or the last individual composition.
\begin{exi}
\begin{figure}
\centering
\input{file_sync.tex}
\caption{A made-up protocol for file synchronisation}
\label{fig:pnls}
\end{figure}
The automaton on Figure \ref{fig:pnls} is a sample protocol for file synchronisation.
The two holes are \(net\) a process to manage network communications, and \(hash\) a process to check files integrity.
The state \(i\) is the initial state.
In that state we want to synchronise the files before working, or work on the local version as a fallback.
If the preferred version is the remote one, we go in state \(d1\) asking the netwotk manager to fetch the lastest version, then ask the integrity checker to validate the files.
If the preferred version is the local one, we go in state \(d2\) asking for the integrity checker to validate the local files, then ask the network manager to push them.
In fallback we use the local version.

We can plug an inactive process for the network manager and the file integrity checker by composing a deadlock automaton in their holes, but always both at the same time.
The reason for that is the simultaneous composition is non-locking, but any individual composition is locking because the state \(d1\) or \(d2\) (depending on the order) becomes a deadlock.
\end{exi}
From this point and in the previous definitions, composition will only refer to non-locking composition.

@@

\section{Refinement relations in restricted cases}\label{sec:prelref}
The main goal of this section is to explain the different aspects of refinement in restricted cases so that they are not all introduced at once in the real refinement relation.
In particular if we have a family of open automata \(A_i \defobject \OAg[i]\), \(0 \leq i \leq n\), let \(A \defobject \OAg\) be such that \(A = A_0\psubst{j_i \mapsto A_i \middle| 1 \leq i \leq n}\).
Recall that there is no relation between the holes of \(A\) and \(A_0\) because \(J = \mpar{\biguplus_{0 \leq i \leq n} J_i} \setminus \mset{j_i}{1 \leq i \leq n}\) is neither subset nor superset of \(J_0\).
So looking at restricted cases where holes are related in a specific manner can help understanding the general case.
Hopefully the intermediate refinement relation introduced can be used as simpler (and less expensive to compute) versions of the real refinement relation in the case the manipulated automata respect some constraints.

We restrict the holes of the related automata so that we can explain the different refinements without their interactions.
In the following definition, \(\triangle \in \mbrc{\subseteq, =, \supseteq}\).
\begin{defi}[Hole-\(\triangle\) simulation]
For two open automata \(A_1 \defobject \OAg[1]\) and \(A_2 \defobject \OAg[2]\), the relation on configurations \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\) is a hole-\(\triangle\) simulation of \(A_1\) by \(A_2\) if:
\defitem \(J_1 \triangle J_2\)
\defitem \(\sigma_{01} \uplus \sigma_{02} \vdash R\mpar{s_{01}, s_{02}}\)
\defitem \(\forall \mpar{s_1, s_2} \in S_1 \times S_2,\)\vspace{-8pt}
\noindent\begin{multline*}
	\mpar{\everymath{\displaystyle}\begin{array}{l}
		\bigsymb{\forall} t_1 \defobject \OTx{1}{}{1}{1} \in \fOT{s_1}, \bigsymb{\exists} \mpar{t_{2x} \defobject \OTx{2}{x}{2x}{2x} \in \fOT{s_2}}^{x \in X}, \\
		\quad \mpar{\forall x \in X, J'_{2x} \cap J_1 = J'_1 \cap J_2} \\[-10pt]
		\nwedge V_1 \uplus V_2 \uplus \fvars{t_1} \vdash R\mpar{s_1, s_2} \wedge g_1 \implies \operatorname*{\bigsymb{\bigvee}}_{x \in X} \mpar{\begin{array}{l}
			\alpha_1 = \alpha_{2x} \wedge \bigwedge_\subbox{j \in J'_{2x} \cap J_1} \beta_{1j} = \beta_{2xj} \\[12pt]
			\nwedge g_{2x} \wedge R\mpar{s'_1, s'_{2x}}\psubst{\psi_1 \uplus \psi_{2x}}
		\end{array}}
	\end{array}} \\
	\wedge V_1 \uplus V_2 \uplus \biguplus_\subbox{t_2 \in \fOT{s_2}} \fvars{t_2} \vdash R\mpar{s_1, s_2} \wedge \bigvee_\subbox{t_2 \in \fOT{s_2}} \fguard{t_2} \implies \bigvee_\subbox{t_1 \in \fOT{s_1}} \fguard{t_1}
\end{multline*}
\end{defi}
The third item of the definition has two parts:
The first part is the main requirement, a (not equivalent) reordering of the pre-simulation requirement so that it better fits SMT.
Instead of matching an effective transition to another effective transition as in the pre-simulation definition, we match an open transition to a family of covering open-transitions.
The second part is directly the deadlock prevention requirement.

@@

\[ \bigsymb{\iff} \]

@@

	\end{array} \hyp{0} \\
